Non-word repetition in Yélî Dnye

### 2021-09-01 getting_frequency_corpus.Rmd
one third through r2

Also, this reveals my rewrites were incomplete, so I added some lines for the onsets that were missing in my initial correspondance list.

The correspondances here looks similar to the one in wrangling but it has more entries because we want to separate each phoneme, and we want to distinguish between short and long

>>>>

Also, this reveals my rewrites were incomplete, so I added some lines for the onsets that were missing in my initial correspondence list.

The correspondences here looks similar to the one in wrangling but it has more entries because we want to separate each phoneme, and we want to distinguish between short and long

### 2021-09-01 manuscript.Rmd
one third through r2

abstract: >
   In non-word repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible. NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns. In this study, we report on NWR results among children (N= 40, aged 3-10 years) learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea. Results make three contributions that are specific, and a fourth that is general. First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects. Second, we documented rather weak effects of item length. Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order. Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations,  and the extent to which they shed light on the factors universally structuring variation  in phonological development at a global and individual level.

@@

knitr::opts_chunk$set(warning = FALSE, message = FALSE, cache = FALSE, fig.pos = "T")
require("papaja")
library(kableExtra)
library(gdata)
library(tidyverse)
library(lme4)
library(openxlsx)
library(sjPlot)

@@

NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@chiat2015non;@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language;@chiat2015nonword;@armon2015assessing].  Non-words can be designed to try to isolate more narrowly certain skills; for instance, one can choose non-words that contain real morphemes in order to load more on prior language experience, or non-words that are shorter to avoid loading on working memory [see a discussion in @chiat2015nonword]. Broadly, however, NWR scores will necessarily reflect to a certain extent long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).

@@

It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture -- a hypothesis discussed for instance in @chiat2015non (pp. 7-8; see also p. 5). In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer. A non-systematic meta-analyses does not provide overwhelming support for this hypothesis [@cristia2021supplementary; SM1].

Nonetheless, given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic analyses.

@@

The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021nonword]. In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education. Based on previous work, we looked at potential increases with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]. Prior research typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool]. Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [e.g., @havron2019effect], and therefore we report on these too.

@@

The fourth research goal we pursued is to use NWR with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size]. Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel; @maddieson2013consonant; with no exceptions to our knowledge].

@@

In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, better than rhyme awareness for instance [e.g., @gathercole1991differentiating]. Therefore, it may not be a general index of, for instance, phonological development, but reflect only certain non-universal language skills. Indeed, @cristia2020infant present their task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations. We return to the question of what was measured here in the Discussion.

@@

Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions. This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, who tend to involve English participants.

@@

Before going into the details of our study design, we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island. As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethnographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible. Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular. International phone calls and digital exchanges that require significant data transfer are typically not an option. Data collection is therefore typically limited to the duration of the researchers' on-island visits.

@@

Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are disyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent dictionary at the time of writing; @levinsonYDgrammar]. The vast majority of syllables use a CV format. A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., "ndap" /ṇṭæp/ *Spondylus shell*) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., "ndapî" /'ṇṭæ.pɯ/). There are also a handful of words starting with /æ/ (e.g., "ala" /æ.'læ/ *here*) and a small collection of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).

@@

@@

Some aspects of the community are relevant for contextualizing our study design and results, particularly regarding sources of individual variation. Specifically, we investigated potential effects of age, gender, maternal education, and birth order. There is nothing particular to note regarding age and gender, but we have some comments that pertain to the other two factors.

@@

This study was approved as part of a larger research effort by the second author. The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041). As discussed in subsection "The Yélî community", the combination of collective child guardianship practices and common billeting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents. Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood [@brownIPchildrearing]. Participation was voluntary; children were invited to participate following indication of approval from an adult caregiver. Regardless of whether they completed the task, children were given a small snack as compensation. Children who showed initial interest but then decided not to participate were also given the snack.

@@

@@

Additionally, we estimated frequency of the phones present  in the target items in a corpus of child-centered recordings [@casillas2020early] by counting the number of word types in which they occurred, and applied the natural logarithm.[^1] Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).

@@

We fitted the child with a headset microphone (Shure SM10A or WH20 XLR with a dynamic microphone on a headband, most children using the former) that fed into the left channel of a Tascam DR40x digital audio recorder. The headsets were designed for adult use and could not be comfortably seated on many children's heads without a more involved adjustment period. To minimize adjustment time, which was uncomfortable for some children given the proximity of the foreign experimenter and equipment, we placed the headband on children's shoulders in these cases, carefully adjusting the microphone's placement so that it was still close to the child's mouth. A research assistant who spoke Yélî Dnye natively sat next to the child throughout the task to provide instructions and, if needed, encouragement. The research assistant coached the child throughout the task to make sure that they understood what they were expected to do. An experimenter (the first author) delivered the pre-recorded stimuli to the research assistant and the child over headphones.

@@

The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software [@Praat], then ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to .wav clips. The assistant then listened through all these paired target-repetition clips randomized across children and repetitions, grouped such that all the clips of the same target were listened to in succession. For each clip, the assistant indicated in a notebook whether the child production was a correct or incorrect repetition and orthographically transcribed the production, noting when the child uttered a recognizable word or phrase and adding the translation equivalent of that word/phrase into English. The assistant was also provided with some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets.

@@

We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequency could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.  Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies [r(`r freq_cor_phone$parameter`)=`r round(freq_cor_phone$estimate,3)`, p `r pval.display(round(freq_cor_phone$p.value,3))`]; and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies [r(`r freq_cor_item$parameter`)=`r round(freq_cor_item$estimate,3)`, p `r pval.display(round(freq_cor_item$p.value,3))`]. Moreover, averaging across participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r(`r cor_ty_nwr$parameter`)=`r printr(cor_ty_nwr$estimate)`, p `r pval.display(round(cor_ty_nwr$p.value,3))`. Therefore, we fit another mixed logistic regression, this time declaring as fixed effects both scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age. As before, the model contained random slopes for both child ID and target. In this model, both cross-linguistic phone frequency (ß = `r both_coeff["avg_fr_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_s","Std. Error"]`, p `r pval.display(both_coeff["avg_fr_s","Pr(>|z|)"])`) and age (ß = `r both_coeff["age.rounded_s","Estimate"]`, SE ß = `r both_coeff["age.rounded_s","Std. Error"]`, p `r pval.display(both_coeff["age.rounded_s","Pr(>|z|)"])`) were significant predictors of whole-item  NWR scores, but corpus phone frequency (ß = `r both_coeff["avg_fr_cor_ty_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_cor_ty_s","Std. Error"]`, p `r pval.display(both_coeff["avg_fr_cor_ty_s","Pr(>|z|)"])`) was not.

@@

In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent]  on NWR scores. We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally. In fact, maternal education correlated with NWR score at about r~.1, which is small. Similarly, NWR scores may not vary greatly with participant gender according to previous work [@chiat2007preschool], and for that as well we find effects of about that size.

@@

WHAT IS BEING MEASURED?
We then added a mention in the discussion that our work does not allow us to tease apart lexical knowledge from phonological knowledge and from working memory in ways that have been done in previous research.

@@

Broadening our discussion to all of the studies in our literature review, we notice that there is rather wide variation of the range of NWR scores found across these samples, and that, in fact, the strength of age effects also varies. We performed some exploratory analyses to see whether features of the languages children were learning could be related to their overall NWR scores. We extracted the number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be longer or shorter based on information in the papers or other sources. Neither of these two predictors explained variance in Figure \@ref(fig:fig-prevlit-overall). It is possible that average word length plays a role, but often researchers incorporate this into their design by including longer items when the native language allows this, with e.g. Sotho non-words having 4-7 syllables in length. To be more certain whether language characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid. We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal. An alternative may be to find ways to regress out some of these effects, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items. As for different strengths of age effects, here as well we are uncertain to what they may be due, but we do hope that these intriguing observations will lead others to collect and share NWR data.

@@

While NWR can, in theory, be used to test a variety of questions about phonological development in any language, previous work has been primarily limited to a handful of related languages spoken in urban, industrialized contexts. The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant gender, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study. Because these latter predictors strongly relate to other language outcomes, the present findings raise many questions, including:  Why do NWR scores would pattern differently across samples? What does that tell us about the relationship between lexical development, phonological development, and the input environment? What is implied about the joint applicability of these outcome measures as a diagnostic indicator for language delays and disorders? While answers to these questions are sought, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a potential association between individual learners' NWR and cross-linguistic phone frequency.

>>>>

abstract: >
   In non-word repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible. NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns. In this study, we report on NWR results among children (N= 40, aged 3-10 years) learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea. Results make three contributions that are specific, and a fourth that is general. First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects. Second, we documented rather weak effects of item length. Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with child sex, maternal education, and birth order. Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations,  and the extent to which they shed light on the factors universally structuring variation  in phonological development at a global and individual level.

@@

knitr::opts_chunk$set(warning = FALSE, message = FALSE, cache = FALSE, fig.pos = "T")
require("papaja")
library(kableExtra)
library(gdata)
library(tidyverse)
library(lme4)
library(openxlsx)
library(sjPlot)
library(stringr)

@@

NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@chiat2015non;@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe in the context of diagnosing language impairments among bilingual children [@meir2016disentangling;@is08042009language;@chiat2015nonword;@armon2015assessing].  Non-words can be designed to try to isolate more narrowly certain skills; for instance, one can choose non-words that contain real morphemes in order to load more on prior language experience, or non-words that are shorter to avoid loading on working memory [see a discussion in @chiat2015nonword]. Broadly, however, NWR scores will necessarily reflect to a certain extent phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).

@@

It is possible that differences are due to language-specific characteristics, including the most common length of words in the lexicon and/or in child-experienced speech in that culture -- a hypothesis discussed for instance in @chiat2015non (pp. 7-8; see also p. 5). In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words tend to be longer. A non-systematic meta-analyses does not provide overwhelming support for this hypothesis [@cristia2021supplementary; SM1].

Nonetheless, given the paucity of research looking at this question, and the diversity of current results, we did not approach this issue within a hypothesis-testing framework but sought instead to provide additional data  on the question, which may be re-used in future meta- or mega-analytic analyses.

@@

The third research area we contribute data to relates to the possibility that children differ from each other in NWR scores in systematic ways. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021nonword]. In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education. Our focus on age stems from previous work [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]. Prior research typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child sex [@chiat2007preschool]. Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [e.g., @havron2019effect], and therefore we report on these too.

@@

The fourth research goal we pursued is to use NWR with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size]. Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant]. To our knowledge, it has never been used with speakers of languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel; @maddieson2013consonant].

@@

In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading; for instance, better than rhyme awareness  [e.g., @gathercole1991differentiating]. Therefore, it may not be a general index of, for instance, phonological development, but reflect only certain non-universal language skills. Indeed, @cristia2020infant present their task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations. We return to the question of what was measured here in the Discussion.

@@

Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions. This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, which involved English participants.

@@

Before going into the details of our study design, we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island. As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethnographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites. Rossel Island lies 250 nautical miles off the coast of mainland PNG and is surrounded by a barrier reef. As a result, transport to and from the island is both infrequent and irregular. International phone calls and digital exchanges that require significant data transfer are typically not an option. Data collection is therefore typically limited to the duration of the researchers' on-island visits.

@@

```{r echo=F}
scan("words_corpus.txt",what="char")->wds

un=unique(wds)
un1=gsub("[:aeiouâêîáóéú]*", " ",un)
count_syll=str_count(un1," ")-1
count_syll[count_syll==0]<-1
round(table(count_syll)/length(count_syll)*100)

Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are disyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent dictionary at the time of writing; @levinsonYDgrammar]. We also approximated word length based on the number of vowel sequences in words found in a conversational corpus (see Stimuli section for details), which suggested a more balanced distribution: 15% monosyllabic, 39% disyllabic, 29% trisyllabic,  and the remaining 17% being longer than that. The vast majority of syllables use a CV format. A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., "ndap" /ṇṭæp/ *Spondylus shell*) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., "ndapî" /'ṇṭæ.pɯ/). There are also a handful of words starting with /æ/ (e.g., "ala" /æ.'læ/ *here*) and a small collection of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).

@@

Before closing this section, it bears mentioning that the language has an established orthography, which includes distinct graphemes for all the contrasts on which our items are based. Although some children in our sample will have started school, at present reading and writing instruction is done only in English (other than writing one's name). This was probably not the case for the majority of mothers of the children in our sample, who will have learned to read and write in Yélî Dnye during their first three years at school. This change from Yélî-initial instruction to English only occurred about 15 years ago. It is possible that there is also some home teaching of Yélî reading and writing, notably for reading the bible.

@@

Some aspects of the community are relevant for contextualizing our study design and results, particularly regarding sources of individual variation. Specifically, we investigated potential effects of age, child sex, maternal education, and birth order. There is nothing particular to note regarding age and child sex, but we have some comments that pertain to the other two factors.

@@

This study was approved as part of a larger research effort by the second author. The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041), including the use of verbal (not written) consent. As discussed in subsection "The Yélî community", the combination of collective child guardianship practices and common billeting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents. Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood [@brownIPchildrearing]. Participation was voluntary; children were invited to participate following indication of approval from an adult caregiver. Regardless of whether they completed the task, children were given a small snack as compensation. Children who showed initial interest but then decided not to participate were also given the snack.

@@

#### Cross-linguistic frequency

@@

#### Within-language frequency

Additionally, we estimated frequency of the phones present  in the target items in a corpus of child-centered recordings [@casillas2020early]. That corpus was constituted by sampling from audio-recordings (7-9 hours long), collected as 10 children aged between 1 month and 3 years went about their day.  The researchers selected 9 2.5-minute clips randomly and 11 5-minute clips by hand (selected to represent peak turn-taking and child vocal activity). These clips were  segmented and transcribed by the lead researcher and a highly knowledgeable local assistant, who speaks Yélî Dnye natively, has ample experience in this kind of research, and often knew all the people talking personally. For more details, please refer to @casillas2020early.

For the present study, we extracted the transcriptions of adult speech (i.e., removing key child and other children's speech) and split them into words using white space. We then removed all English and Pidgin words. The resulting corpus contained a total of 18,934 word tokens of 1,686 unique word types. To get our phone frequency measure, we counted the number of word types in which the phone occurred, and applied the natural logarithm.[^1] Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).

@@

We fitted the child with a headset microphone (Shure SM10A or WH20 XLR with a dynamic microphone on a headband, most children using the former) that fed into the left channel of a Tascam DR40x digital audio recorder. The headsets were designed for adult use and could not be comfortably seated on many children's heads without a more involved adjustment period. To minimize adjustment time, which was uncomfortable for some children given the proximity of the foreign experimenter and equipment, we placed the headband on children's shoulders in these cases, carefully adjusting the microphone's placement so that it was still close to the child's mouth. A research assistant who spoke Yélî Dnye natively, and who could also hear the instructions over headphones, sat next to the child throughout the task to provide instructions and, if needed, encouragement. The research assistant coached the child throughout the task to make sure that they understood what they were expected to do. Finally, an experimenter (the first author) was also fitted with headphones and a microphone; she was in charge of delivering the pre-recorded stimuli to the research assistant, the child, and herself over headphones.

@@

The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software [@Praat], then ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to .wav clips. The assistant then listened through all these paired target-repetition clips randomized across children and repetitions, grouped such that all the clips of the same target were listened to in succession. For each clip, the assistant indicated in a notebook whether the child production was a correct or incorrect repetition and orthographically transcribed the production, noting when the child uttered a recognizable word or phrase and adding the translation equivalent of that word/phrase into English. The assistant was also provided with some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets. After completing all of her annotations, she double-checked them by listening to them twice.

@@

### NWR and typology: NWR as a function of within-language phone frequency

We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequency could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.

Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies [r(`r freq_cor_phone$parameter`)=`r round(freq_cor_phone$estimate,3)`, p `r pval.display(round(freq_cor_phone$p.value,3))`]; and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies [r(`r freq_cor_item$parameter`)=`r round(freq_cor_item$estimate,3)`, p `r pval.display(round(freq_cor_item$p.value,3))`]. Moreover, averaging across participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r(`r cor_ty_nwr$parameter`)=`r printr(cor_ty_nwr$estimate)`, p `r pval.display(round(cor_ty_nwr$p.value,3))`. Therefore, we fit another mixed logistic regression, this time declaring as fixed effects both scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age. As before, the model contained random slopes for both child ID and target. In this model, both cross-linguistic phone frequency (ß = `r both_coeff["avg_fr_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_s","Std. Error"]`, p `r pval.display(both_coeff["avg_fr_s","Pr(>|z|)"])`) and age (ß = `r both_coeff["age.rounded_s","Estimate"]`, SE ß = `r both_coeff["age.rounded_s","Std. Error"]`, p `r pval.display(both_coeff["age.rounded_s","Pr(>|z|)"])`) were significant predictors of whole-item  NWR scores, but corpus phone frequency (ß = `r both_coeff["avg_fr_cor_ty_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_cor_ty_s","Std. Error"]`, p `r pval.display(both_coeff["avg_fr_cor_ty_s","Pr(>|z|)"])`) was not.

@@

In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent]  on NWR scores. We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally. In fact, maternal education correlated with NWR score at about r~.1, which is small. Similarly, NWR scores may not vary greatly with participant sex according to previous work [@chiat2007preschool], and for that as well we find effects of about that size.

@@

@@

Broadening our discussion to all of the studies in our literature review, we notice that there is rather wide variation of the range of NWR scores found across these samples, and that, in fact, the strength of age effects also varies. We performed some exploratory analyses to see whether features of the languages children were learning could be related to their overall NWR scores. We extracted the number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be longer or shorter based on information in the papers or other sources. Neither of these two predictors explained variance in Figure \@ref(fig:fig-prevlit-overall). It is possible that average word length plays a role, but often researchers incorporate this into their design by including longer items when the native language allows this, with e.g. Sotho non-words having 4-7 syllables in length. To be more certain whether language-specific characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid. We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal. An alternative may be to find ways to regress out some of these effects, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items. As for different strengths of age effects, here as well we are uncertain to what they may be due, but we do hope that these intriguing observations will lead others to collect and share NWR data.

### Limitations

Before closing, we would like to point out some salient limitations of the current work. To begin with, we only employed one set of non-words, in which not all characteristics that previous work suggest matter were manipulated [@chiat2015non]. As a result, we only have a rather whole-sale measure of performance, and we do not know to what extent lexical knowledge, pure phonological knowledge, and working memory, among others, contribute to children's performance. Similarly, our items varied systematically in length and typological frequency of the sounds included, but not in other potential dimensions (such as whether the items contained morphemes of the language or not). Additionally, we only had a single person interacting with children as well as interpreting children's production, so we do not know to what extent our findings generalize to other experimenters and research assistants. In addition, since both stimuli presentation and production data collected were audio-only, neither the children nor our research assistant were able to integrate visual cues in their interpretation. Although we know from other work that adults' perceptual performance on these types of sounds is well above chance from audio-only presentation [REF],  language processing for the majority of children will be audiovisual in natural conditions, and thus it may be interesting in the future to capture this aspect of speech.

@@

While NWR can, in theory, be used to test a variety of questions about phonological development in any language, previous work has been primarily limited to a handful of related languages spoken in urban, industrialized contexts. The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant sex, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study. Because these latter predictors strongly relate to other language outcomes, the present findings raise many questions, including:  Why do NWR scores would pattern differently across samples? What does that tell us about the relationship between lexical development, phonological development, and the input environment? What is implied about the joint applicability of these outcome measures as a diagnostic indicator for language delays and disorders? While answers to these questions are sought, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a potential association between individual learners' NWR and cross-linguistic phone frequency.
